Summary of the things that I learned

### 2024-10-14 100-introduction.tex
Slightly improve the introduction after feedback

However, because the computation part of such a proof involves a lot of code, and computer code tends to contain bugs, accepting a proof by computer involves a certain amount of trust. Therefore, both of the theorems mentioned have subsequently also been proved using `computer proof assistants'\autocite{formalized-four-color-theorem}\autocite{formalized-Kepler-conjecture}. Such proof assistants, like `rocq', `lean' and `Agda', are computer programs with only a very small `trusted codebase'\footnote{Trusted codebase means the part of the codebase of which we hope that it is correct. The rest of the codebase is checked using this trusted core, so if the trusted core is correct, the rest is too.}, that can verify mathematical reasoning. In this way, if we trust the small core of such a proof assistant, and the proof assistant says that a proof is correct, then we can trust that indeed, the proof is correct.

@@

Now, much of contemporary mathematics is built on the foundation of set theory, which we will often refer to as \iindex{classical mathematics}. Usually, `set theory' refers to ZFC: Zermelo Fraenkel set theory with the axiom of choice, developed in the early 1900s. On the other hand, most of the formalization in proof assistants is built on the foundation of type theory. This is because computers can reason better with types than with sets, as evidenced by the many typed programming languages that are around nowadays. However, type theory has been around longer than computers have: it was initially developed by Bertrand Russell in the early 1900s, to create a foundation of mathematics that avoided Russell's paradox\footnote{``Consider the set of all sets that do not contain themself. Does this set contain itself?''}.

Nowadays, there are many flavours of type theory around. The proof assistant rocq is based on the `calculus of constructions' whereas the proof assistant Agda works with the `unified theory of dependent types'. Every flavour has its advantages and disadvantages, but in this thesis we will work with `univalent foundations', set forth by Vladimir Voevodsky \autocite{voevodsky-univalent-foundations}. In univalent foundations, the common mathematical principle that `things with a similar structure\footnote{For example, two sets with a bijection between them have a similar structure. Also, two graphs $ G $ and $ H $ have a similar structure when there is a bijection $ f $ between their sets of vertices, such that for every two vertices $ v, w : G $, we have a bijection between the set of edges from $ v $ to $ w $, and the set of edges from $ f(v) $ to $ f(w) $.}, are the same' is made explicit. The principle is used often in modern mathematics, and making this explicit in the type theory that we work with, allows us to use the principle when formalizing.

However, note that two objects can often have a similar structure in more than one way. For example, the set $ \{ \top, \bot \} $ has two bijections to the set $ \{ 0, 1 \} $. To accomodate for this, in univalent foundations, two things can sometimes `be the same' in more way than one. For example, we have two equalities between $ \{ \top, \bot \} $ and $ \{ 0, 1 \} $. This means that the concept of `sameness' is more intricate to work with in univalent foundations than it is in set theory. This is one of the reasons that there are subtleties involved in transferring definitions and proofs from set theory to univalent foundations. It is therefore still a topic of ongoing research how well different parts of set-based mathematics can be transferred to univalent foundations and what subtleties pop up when material from set-based mathematics is transferred to univalent foundations. For example, often in set-based mathematics, two definitions are equivalent if we assume the axiom of choice. However, sometimes in univalent foundations, the axiom of choice is not sufficient to make two such definitions equivalent, and then we have two different definitions. It is then interesting to explore the behaviour of these two different definitions in univalent foundations.

In this thesis, we study the paper `classical lambda calculus in modern dress' through the lens of univalent foundations, work out the details of the proofs and formalize part of the paper and its preliminaries. The major contributions are
\begin{itemize}
  \item The category theoretical preliminaries needed to understand the paper (Chapter \ref{ch:category-theory}), as well as the work of Hyland's predecessors that he expands upon (Chapter \ref{ch:previous-work}), and more detailed versions of his definitions and proofs, complemented with some examples (Chapters \ref{ch:algebraic-structures} and \ref{ch:the-paper}). This makes Hyland's paper more accessible to computer scientists.
  \item The translation of Hyland's paper and the work of his predecessors to univalent foundations (Chapter \ref{ch:the-paper}). This shows that his work can be translated to univalent foundations, and contributes to the knowledge about translating classical mathematics to univalent foundations in general.
  \item A formalization of part of the paper in rocq (Chapter \ref{ch:the-formalization}). This contributes to the knowledge about formalization in general.
  \item A new proof for the fundamental theorem of the $ \lambda $-calculus (Section \ref{sec:elementary-fundamental-theorem}). This version is easier to read and verify, because it is more elementary and uses less category theory than Hyland's proof.
  \item An analysis of the behaviour of the Karoubi envelope in univalent foundations in general (Section \ref{sec:karoubi-envelope}), and in the specific case of Paul Taylor's and Martin Hyland's work (Remark \ref{rem:difference-Taylor-Hyland}). Here we see an example of two definitions that are equivalent in classical mathematics, but become different in univalent foundations. It is interesting to see the differences between the behaviour of the two definitions in univalent foundations.
\end{itemize}

% 600-paper.tex
Now, most of this thesis works towards Chapter \ref{ch:the-paper}, about the three main proofs of Hyland's paper.
% 300-univalent-foundations.tex
Because this thesis works from a univalent point of view, Chapter \ref{ch:univalent-foundations} introduces univalent foundations and builds the preliminary knowledge in that area for the rest of the paper.
% 500-previous-work.tex
As mentioned before, the work of Hyland's predecessors that is covered in Chapter \ref{ch:previous-work} helps to understand his paper.
% 200-category-theory.tex
The category theoretical preliminaries for understanding Hyland's paper are covered in Chapter \ref{ch:category-theory}. It is probably wise to skim this chapter, and periodically come back to it to better understand the material in the other chapters.
% 400-algebraic-structures.tex
Hyland's three main theorems deal with a couple of objects that he introduces. These objects, together with some properties and examples, are introduced in Chapter \ref{ch:algebraic-structures}.
% 700-formalization.tex
Lastly, the formalization of part of the material is discussed in Chapter \ref{ch:the-formalization}.

>>>>

However, because the computation part of such a proof involves a lot of code, and computer code tends to contain bugs, accepting a proof by computer involves a certain amount of trust. Therefore, both of the theorems mentioned have subsequently also been proved using `computer proof assistants' \autocite{formalized-four-color-theorem}\autocite{formalized-Kepler-conjecture}. Such proof assistants, like `rocq', `lean' and `Agda', are computer programs with only a very small `trusted codebase'\footnote{The trusted codebase is the part of the codebase of which we have to trust that it is correct. The rest of the codebase is checked using this trusted core, so if the trusted core is correct, the rest is too.}, that can verify mathematical reasoning. In this way, if we trust the small core of such a proof assistant, and the proof assistant says that a proof is correct, then we can trust that indeed, the proof is correct.

@@

Now, much of contemporary mathematics is built on the foundation of set theory, which we will often refer to as \iindex{classical mathematics}. Usually, `set theory' refers to ZFC: Zermelo Fraenkel set theory with the axiom of choice, developed in the early 1900s \autocite{Zermelo-set-theory}. On the other hand, most of the formalization in proof assistants is built on the foundation of type theory. This is because computers can reason better with types than with sets, as evidenced by the many typed programming languages that are around nowadays. However, type theory has been around longer than computers have: it was initially developed by Bertrand Russell in the early 1900s (\autocite{Russell-type-theory}, Appendix B), to create a foundation of mathematics that avoided Russell's paradox\footnote{Russell's paradox is the question ``Consider the set of all sets that do not contain themself. Does this set contain itself?'' Answering this question with either `yes' or `no' leads to a contradiction.}.

Nowadays, there are many flavours of type theory around. The proof assistant rocq is based on the `calculus of constructions' whereas the proof assistant Agda works with the `unified theory of dependent types'. Every flavour has its advantages and disadvantages, but in this thesis we will work with `univalent foundations', set forth by Vladimir Voevodsky \autocite{voevodsky-univalent-foundations}. The advantage of working with univalent foundations is that it makes explicit the common mathematical principle that `things with a similar structure\footnote{For example, two sets with a bijection between them have a similar structure. Also, two graphs $ G $ and $ H $ have a similar structure when there is a bijection $ f $ between their sets of vertices, such that for every two vertices $ v, w : G $, we have a bijection between the set of edges from $ v $ to $ w $, and the set of edges from $ f(v) $ to $ f(w) $.} are the same', which allows us to use it when formalizing.

Unfortunately, there is a trade-off here: we do not get `similar objects are the same' for free. This is because two objects can often have a similar structure in more than one way. For example, the set $ \{ \top, \bot \} $ has two bijections to the set $ \{ 0, 1 \} $. To accomodate for this, in univalent foundations, two things can sometimes `be the same' in more ways than one. For example, we have two equalities between $ \{ \top, \bot \} $ and $ \{ 0, 1 \} $. This means that the concept of `sameness' becomes more complicated to work with in univalent foundations than it is in set theory. This is one of the reasons that there are subtleties involved in transferring definitions and proofs from set theory to univalent foundations. It is therefore still a topic of ongoing research how well different parts of set-based mathematics can be transferred to univalent foundations and what subtleties pop up when material from set-based mathematics is transferred to univalent foundations. For example, often in set-based mathematics, two definitions are equivalent if we assume the axiom of choice. However, sometimes in univalent foundations, the axiom of choice is not sufficient to make two such definitions equivalent, and then we have two different definitions. It is then interesting to explore the behaviour of these two different definitions in univalent foundations.

In this thesis, we study the paper `classical lambda calculus in modern dress' through the lens of univalent foundations, work out the details of the proofs, and formalize part of the paper and its preliminaries. The major contributions are:
\begin{itemize}
  \item More detailed versions of Hyland's definitions and proofs, complemented with some examples (Chapters \ref{ch:algebraic-structures} and \ref{ch:the-paper}). Also, the work of Hyland's predecessors that he expands upon (Chapter \ref{ch:previous-work}), and most of the category theoretical preliminaries needed to understand the paper (Chapter \ref{ch:category-theory}), to make this thesis reasonably self-contained. All this makes Hyland's paper more accessible to computer scientists.
  \item The translation of Hyland's paper and the work of his predecessors to univalent foundations (Chapter \ref{ch:the-paper}). This shows that his work can be translated to univalent foundations, and contributes to the knowledge about translating classical mathematics to univalent foundations in general.
  \item A formalization of part of the paper in rocq (Chapter \ref{ch:the-formalization}). This contributes to the knowledge about formalization in general.
  \item A new proof for the fundamental theorem of the $ \lambda $-calculus (Section \ref{sec:elementary-fundamental-theorem}). This version is easier to read and verify, because it is more elementary and uses less category theory than Hyland's proof.
  \item An analysis of the behaviour of the Karoubi envelope in univalent foundations in general (Section \ref{sec:karoubi-envelope}), and in the specific case of Paul Taylor's and Martin Hyland's work (Remark \ref{rem:difference-Taylor-Hyland}). Here we see an example of two definitions that are equivalent in classical mathematics, but become different in univalent foundations. It is interesting to see the differences between the behaviour of the two definitions in univalent foundations.
\end{itemize}

% 600-paper.tex
Now, most of this thesis works towards Chapter \ref{ch:the-paper}, about the three main proofs of Hyland's paper.
% 300-univalent-foundations.tex
Because this thesis works from a univalent point of view, Chapter \ref{ch:univalent-foundations} introduces univalent foundations and builds the preliminary knowledge in that area for the rest of the paper.
% 500-previous-work.tex
As mentioned before, the work of Hyland's predecessors that is covered in Chapter \ref{ch:previous-work} helps to understand his paper.
% 200-category-theory.tex
The category theoretical preliminaries for understanding Hyland's paper are covered in Chapter \ref{ch:category-theory}. It is probably wise to skim this chapter, and periodically come back to it to better understand the material in the other chapters.
% 400-algebraic-structures.tex
Hyland's three main theorems deal with a couple of objects that he introduces. These objects, together with some properties and examples, are introduced in Chapter \ref{ch:algebraic-structures}.
% 700-formalization.tex
Lastly, the formalization of part of the material covered in this thesis is discussed in Chapter \ref{ch:the-formalization}.

### 2024-10-14 500-previous-work.tex
Add links back from the conclusion
Slightly improve the introduction after feedback

  During the 1960s, computer scientists sought for nontrivial examples of reflexive objects, there is a quote by Dana Scott that ``Lambda-calculus has no mathematical models!''\autocite{strachey}. However, in 1969, the same Dana Scott discovered that the category of (continuous) lattices with (Scott-)continuous functions between them has a nontrivial reflexive object, with the retraction onto the function spaces being even an isomorphism. Such a reflexive object $ D_\infty $ is obtained by starting with an arbitrary lattice $ D_0 $, iteratively taking $ D_{n + 1} = D_n^{D_n} $, with a retraction (in the `wrong' direction) $ r: D_n^{D_n} \to D_n $, and then passing to the limit $ D_\infty = \lim_{\leftarrow} D_n $ (for the main result, see Theorem 4.4 in \autocite{scott-continuous}, page 127).
\end{remark}

@@

This brings us to the main theorem of this section (\autocite{taylor}, 5.1.8).
\begin{theorem}
  If we assume the axiom of choice, $ \R $ is cartesian closed, relative to the given class of display maps.
\end{theorem}
\begin{proof}
  Let $ \alpha: B \to A $ be a display map. We have mere existence of some $ X: \R^A $ and an isomorphism $ g: (X, \alpha) \xrightarrow \sim (\sum B, p_X) $. We need to show that there is a right adjoint to the pullback functor $ \alpha^* $. Since under assumption of the axiom of choice, $ \alpha^* $ is defined as a composition of a precomposition functor with two equivalences of categories, this is equivalent to showing that there is a right adjoint to the precomposition functor $ (\alpha \cdot -): \R^A \to \R^B $ (note that $ \alpha \cdot - $ sends $ f $ to $ f \circ \alpha $).

>>>>

  During the 1960s, computer scientists sought for nontrivial examples of reflexive objects, there is a quote by Dana Scott that ``Lambda-calculus has no mathematical models!'' \autocite{strachey}. However, in 1969, the same Dana Scott discovered that the category of (continuous) lattices with (Scott-)continuous functions between them has a nontrivial reflexive object, with the retraction onto the function spaces being even an isomorphism. Such a reflexive object $ D_\infty $ is obtained by starting with an arbitrary lattice $ D_0 $, iteratively taking $ D_{n + 1} = D_n^{D_n} $, with a retraction (in the `wrong' direction) $ r: D_n^{D_n} \to D_n $, and then passing to the limit $ D_\infty = \lim_{\leftarrow} D_n $ (for the main result, see Theorem 4.4 in \autocite{scott-continuous}, page 127).
\end{remark}

@@

This brings us to the main theorem of this section (\autocite{taylor}, 5.1.8).
\begin{theorem}\label{thm:Taylor}
  If we assume the axiom of choice, $ \R $ is cartesian closed, relative to the given class of display maps.
\end{theorem}
\begin{proof}
  Let $ \alpha: B \to A $ be a display map. We have mere existence of some $ X: \R^A $ and an isomorphism $ g: (X, \alpha) \xrightarrow \sim (\sum B, p_X) $. We need to show that there is a right adjoint to the pullback functor $ \alpha^* $. Since under assumption of the axiom of choice, $ \alpha^* $ is defined as a composition of a precomposition functor with two equivalences of categories, this is equivalent to showing that there is a right adjoint to the precomposition functor $ (\alpha \cdot -): \R^A \to \R^B $ (note that $ \alpha \cdot - $ sends $ f $ to $ f \circ \alpha $).

### 2024-10-14 800-conclusion.tex
Slightly improve the introduction after feedback

Lastly, we saw how part of the material in this thesis was formalized, and we evaluated the choices that were made in the formalization (Chapter \ref{ch:the-formalization}). Unfortunately, due to the very time-consuming nature of formalization, not all of the material could be formalized. In future work, it would be interesting to see which version of Hyland's fundamental theorem would lend itself best to formalization. Personally, I would guess it is the most elementary one, exhibited in Section \ref{sec:elementary-fundamental-theorem}.

>>>>

Lastly, we saw how part of the material in this thesis was formalized, and we evaluated the choices that were made in the formalization (Chapter \ref{ch:the-formalization}). For some of the very complicated mathematics with a lot of bookkeeping, like the proofs about the Yoneda embedding or the theory of extensions, this formalization constitutes an additional guarantee that it is correct. In some other instances, the process of formalizing contributed to the realization that a lemma should be stated in more generality (see Section \ref{sec:quotients}). Unfortunately, due to the very time-consuming nature of formalization, not all of the material in Hyland's paper could be formalized. In future work, it would be interesting to see which version of the fundamental theorem would lend itself best to formalization. Personally, I would guess it is the most elementary one, exhibited in Section \ref{sec:elementary-fundamental-theorem}.
